---
title: "Advanced Data Visualization"
author: "Isabelle Garza"
description: "Reflection on advanced ggplot2 techniques — scales, composition, annotation, and the grammar of graphics."
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    code-line-numbers: true
execute:
  warning: false
  error: true
  freeze: auto
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(patchwork)
library(forcats)
library(gt)
```

This page documents reflections on advanced data visualization principles and demonstrates key techniques learned through workshops by Thomas Lin Pedersen, David Robinson, and Cédric Scherer, as well as the `gt` and `gtExtras` table packages.

## The Grammar of Graphics

The workshop by Thomas Lin Pedersen was transformative. The core insight was to treat plots not as monolithic final products, but as **composable objects built from reusable parts**. This philosophy, rooted in the grammar of graphics, demystifies complexity.

> By consciously separating data, aesthetics, geoms, and guides, even the most intricate visualizations become manageable.

## Scales and Transformations

Scales are the critical translators between your data and your audience. A plot's ability to communicate honestly and effectively hinges on scale choices.

```{r}
#| label: scales-demo
#| fig-width: 10
#| fig-height: 4

df <- data.frame(
  category = c("A", "B", "C"),
  value = c(10, 20, 1000)
)

p1 <- ggplot(df, aes(x = category, y = value)) +
  geom_col(fill = "#3498db", alpha = 0.85) +
  labs(title = "Linear Scale", subtitle = "Outlier dominates the view") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p2 <- ggplot(df, aes(x = category, y = value)) +
  geom_col(fill = "#e74c3c", alpha = 0.85) +
  scale_y_log10() +
  labs(title = "Log10 Scale", subtitle = "Relative differences revealed") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p1 + p2
```

The same data tells a completely different story depending on the scale. Linear scales can hide small but meaningful differences when outliers are present, while log scales reveal relative proportions.

## Three Essential Functions

From David Robinson's impromptu data visualization workflow, three functions stood out as essential for rapid exploratory analysis:

### `dplyr::count()` — Quick Frequency Analysis

```{r}
#| label: count-demo
#| fig-width: 8
#| fig-height: 5

mpg %>%
  count(class) %>%
  ggplot(aes(x = reorder(class, n), y = n)) +
  geom_col(fill = "#2ecc71", alpha = 0.85) +
  coord_flip() +
  labs(
    title = "Vehicle Class Distribution",
    subtitle = "count() provides instant frequency summaries",
    x = "Car Class", y = "Count"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

### `tidyr::separate()` — Splitting Combined Fields

```{r}
#| label: separate-demo

messy_df <- data.frame(id = 1:3, info = c("Model_A_Red", "Model_B_Blue", "Model_C_Green"))

clean_df <- messy_df %>%
  separate(info, into = c("prefix", "model", "color"), sep = "_")

clean_df %>%
  gt() %>%
  tab_header(title = "Cleaned Data with separate()")
```

### `forcats::fct_reorder()` — Meaningful Ordering

```{r}
#| label: fct-reorder-demo
#| fig-width: 8
#| fig-height: 5

mpg %>%
  mutate(class = fct_reorder(class, hwy, .fun = median)) %>%
  ggplot(aes(x = class, y = hwy)) +
  geom_boxplot(fill = "#9b59b6", alpha = 0.6) +
  coord_flip() +
  labs(
    title = "Highway MPG by Car Class",
    subtitle = "Ordered by median MPG using fct_reorder() — patterns jump out",
    x = "Car Class", y = "Highway MPG"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

## Tables as Data Products

The `gt` package redefined my perception of tables. A table can be more than a crude printout — it can be a sophisticated data product with its own visual appeal.

```{r}
#| label: gt-demo

mtcars %>%
  group_by(cyl) %>%
  summarize(
    n = n(),
    avg_mpg = round(mean(mpg), 1),
    avg_hp = round(mean(hp), 0),
    avg_wt = round(mean(wt), 2),
    .groups = "drop"
  ) %>%
  gt() %>%
  tab_header(
    title = "Engine Performance Summary",
    subtitle = "Grouped by number of cylinders"
  ) %>%
  cols_label(
    cyl = "Cylinders",
    n = "Count",
    avg_mpg = "Avg MPG",
    avg_hp = "Avg HP",
    avg_wt = "Avg Weight (tons)"
  )
```

> While charts excel at showing patterns, tables are superior for presenting exact values. The combination of `gt`'s formatting rigor and the ability to embed mini visuals means a single table can communicate trend, spread, and precise figures simultaneously.

## Design Principles from Cédric Scherer

Three key principles for intentional, effective data visualization:

1. **Use perceptually uniform palettes** — Palettes like Viridis ensure that changes in color value are perceived as equal changes in data value, without introducing false visual emphasis.

2. **Refine typography and space** — Consistent font sizes, adjusted margins, and lighter gridlines create a clear visual hierarchy that guides the viewer's eye without distraction.

3. **Practice intentional annotation** — Annotate sparingly. Highlight the two or three key data points, and let the rest of the data speak for itself.

## Life Expectancy: Top and Bottom 5 in Asia

Using `patchwork` to compose multiple plots into a single, coherent narrative.

```{r}
#| label: asia-life-exp
#| fig-width: 12
#| fig-height: 5

library(gapminder)

asia_2007 <- gapminder %>%
  filter(year == 2007, continent == "Asia")

top5 <- asia_2007 %>% slice_max(lifeExp, n = 5)
bottom5 <- asia_2007 %>% slice_min(lifeExp, n = 5)

p1 <- ggplot(top5, aes(x = reorder(country, lifeExp), y = lifeExp)) +
  geom_col(fill = "#2ecc71", alpha = 0.85) +
  coord_flip() +
  labs(title = "Top 5 Life Expectancy", x = "", y = "Life Expectancy (years)") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p2 <- ggplot(bottom5, aes(x = reorder(country, lifeExp), y = lifeExp)) +
  geom_col(fill = "#e74c3c", alpha = 0.85) +
  coord_flip() +
  labs(title = "Bottom 5 Life Expectancy", x = "", y = "Life Expectancy (years)") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p1 + p2 +
  plot_annotation(
    title = "Life Expectancy in Asia (2007)",
    subtitle = "Composed with patchwork — comparing extremes side by side",
    theme = theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "gray40")
    )
  )
```

## Reflections on Mastering Data Visualization

**Challenges identified:** Balancing creativity and clarity, managing the many ggplot2 details that govern scales, themes, and layering. Moving beyond simple charts can feel overwhelming.

**Plan to overcome:** Execute every new technique in small loops — keep it simple, add one change, and re-read the message. Maintain a short checklist based on module objectives covering correlations, multi-variable layers, animation, and maps. Recreate one plot from tutorials each week to build confidence through practice.
