---
title: "Data Visualization in R"
author: "Isabelle Garza"
description: "Exploring data visualization techniques using ggplot2, plotly, and the grammar of graphics."
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    code-line-numbers: true
execute:
  warning: false
  error: true
  freeze: auto
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(ggrepel)
library(ggthemes)
library(scales)
library(plotly)
library(GGally)
library(gt)
library(wooldridge)

df <- attend
df <- as_tibble(df)
```

This page showcases data visualization work completed as part of my data science coursework. Each section demonstrates a different visualization technique using R and the `ggplot2` ecosystem, applied to the `attend` dataset from the Wooldridge package.

## Interactive Data Table

Using the `gt` package to create a professional, formatted data table with custom column labels, percentage formatting, and styled borders.

```{r}
#| label: gt-table

df %>%
  head(12) %>%
  gt() %>%
  cols_label(
    termGPA = "Term GPA",
    priGPA = "Prior GPA",
    atndrte = "Attendance Rate",
    hwrte = "Homework Rate",
    frosh = "Freshman",
    soph = "Sophomore"
  ) %>%
  tab_options(
    table.border.top.style = "solid",
    table.border.top.color = "#0b5cad",
    table.border.bottom.style = "solid",
    table.border.bottom.color = "#0b5cad"
  ) %>%
  tab_header(
    title = "Student Attendance Dataset",
    subtitle = "Sample from the Wooldridge 'attend' dataset"
  ) %>%
  fmt_percent(
    columns = atndrte, decimals = 1,
    scale_values = FALSE
  ) %>%
  fmt_number(
    columns = c(termGPA, priGPA), decimals = 2
  )
```

## Scatter Plot: Prior GPA vs. Missed Classes

A faceted scatter plot examining whether the relationship between prior GPA and missed classes differs between sophomores and non-sophomores. The linear regression lines reveal a negative trend — higher GPA is associated with fewer missed classes — and the pattern holds across both groups.

```{r}
#| label: scatter-gpa-missed
#| fig-width: 10
#| fig-height: 5

df_sophomore_recoded <- df %>%
  mutate(soph = factor(ifelse(soph == "1", "Sophomore", "Non-Sophomore")))

df_sophomore_recoded %>%
  ggplot(aes(priGPA, missed)) +
  geom_point(size = 2, color = "#2c3e50", alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#e74c3c", linewidth = 1.2) +
  facet_wrap(~ soph) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Relationship Between Prior GPA and Missed Classes",
    subtitle = "Faceted by sophomore status — higher GPA correlates with fewer absences",
    x = "Prior GPA",
    y = "Number of Missed Classes"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold", size = 12)
  )
```

## ACT Score Distribution

A bar plot showing the distribution of ACT scores in the dataset. The distribution approximates a normal (bell-shaped) curve, which is consistent with standardized test score patterns.

```{r}
#| label: act-distribution
#| fig-width: 8
#| fig-height: 5

ggplot(df, aes(x = ACT)) +
  geom_bar(fill = "#3498db", alpha = 0.85) +
  labs(
    title = "Distribution of ACT Scores",
    subtitle = "Approximately normal distribution across the student sample",
    x = "ACT Score",
    y = "Count"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

## Comparing Bar Chart Types

A critical comparison of dodge, stacked, and faceted bar charts to determine which chart type most accurately represents the relationship between ACT scores, final exam performance, and sophomore status.

::: {.panel-tabset}

### Dodge Chart (Means)

```{r}
#| label: dodge-means
#| fig-width: 10
#| fig-height: 5

df_averages <- df %>%
  mutate(soph_label = factor(soph, levels = c(0, 1),
                             labels = c("Not Sophomore", "Sophomore"))) %>%
  group_by(ACT, soph_label) %>%
  summarise(mean_final = mean(final, na.rm = TRUE), .groups = "drop")

ggplot(df_averages, aes(x = ACT, y = mean_final, fill = soph_label)) +
  geom_col(position = "dodge", width = 0.9) +
  scale_fill_manual(values = c("Not Sophomore" = "#3498db", "Sophomore" = "#e74c3c")) +
  labs(
    title = "Mean Final Score by ACT and Sophomore Status",
    subtitle = "Dodge chart using averaged scores — the most accurate representation",
    x = "ACT Score", y = "Mean Final Score", fill = "Status"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"))
```

### Stacked Chart (Means)

```{r}
#| label: stacked-means
#| fig-width: 10
#| fig-height: 5

ggplot(df_averages, aes(x = ACT, y = mean_final, fill = soph_label)) +
  geom_bar(stat = "identity", width = 0.9) +
  scale_fill_manual(values = c("Not Sophomore" = "#3498db", "Sophomore" = "#e74c3c")) +
  labs(
    title = "Mean Final Score by ACT and Sophomore Status",
    subtitle = "Stacked chart — harder to compare individual group values",
    x = "ACT Score", y = "Mean Final Score", fill = "Status"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

### Faceted Chart (Means)

```{r}
#| label: faceted-means
#| fig-width: 10
#| fig-height: 6

ggplot(df_averages, aes(x = ACT, y = mean_final, fill = soph_label)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  facet_grid(soph_label ~ .) +
  scale_fill_manual(values = c("Not Sophomore" = "#3498db", "Sophomore" = "#e74c3c")) +
  labs(
    title = "Mean Final Score by ACT and Sophomore Status",
    subtitle = "Faceted view — clear separation for individual group analysis",
    x = "ACT Score", y = "Mean Final Score"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    strip.background = element_rect(fill = "#ecf0f1"),
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold")
  )
```

:::

**Key Finding:** The dodge and faceted charts using *averaged* data accurately represent the underlying patterns. The stacked chart makes group comparison difficult. Higher ACT scores consistently predict higher final scores, and sophomores maintain a modest advantage across nearly all ACT levels.

## Boxplot with Jittered Points

A boxplot enhanced with jittered individual data points provides the most comprehensive view — showing the median, spread, outliers, and the full distribution simultaneously.

```{r}
#| label: boxplot-jitter
#| fig-width: 12
#| fig-height: 6

ggplot(df_sophomore_recoded, aes(x = factor(ACT), y = final, fill = soph)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7, width = 0.6) +
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.7),
    aes(color = soph), alpha = 0.5, size = 1.5, show.legend = FALSE
  ) +
  scale_fill_manual(values = c("Non-Sophomore" = "#3498db", "Sophomore" = "#2ecc71")) +
  scale_color_manual(values = c("Non-Sophomore" = "#2c3e50", "Sophomore" = "#27ae60")) +
  labs(
    title = "Final Scores by ACT and Sophomore Status",
    subtitle = "Boxplots reveal distribution spread; jittered points show individual observations",
    x = "ACT Score", y = "Final Score", fill = "Status"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

**Insight:** Sophomores consistently outperform non-sophomores with slightly elevated medians across most ACT levels, suggesting additional academic experience contributes to stronger performance.

## Attendance Rate by ACT Score

Examining how freshman status moderates the relationship between ACT scores and class attendance.

```{r}
#| label: attendance-scatter
#| fig-width: 10
#| fig-height: 5

df_freshman_recoded <- df %>%
  mutate(
    freshman_label = factor(frosh, levels = c(0, 1),
                            labels = c("Non-Freshman", "Freshman"))
  )

ggplot(df_freshman_recoded, aes(x = ACT, y = atndrte)) +
  geom_point(alpha = 0.4, color = "#3498db") +
  geom_smooth(method = "lm", se = FALSE, color = "#e74c3c", linewidth = 1.2) +
  facet_wrap(~ freshman_label) +
  labs(
    title = "Attendance Rate vs. ACT Score",
    subtitle = "Freshmen attend less and show a weaker ACT–attendance relationship",
    x = "ACT Score", y = "Attendance Rate (%)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold", size = 12)
  )
```

**Finding:** Freshmen not only attend less across all ACT levels but also show a diminished response to ACT — the slope for freshmen is notably shallower, indicating that factors beyond academic aptitude drive their attendance patterns.

## Correlation Analysis

A comprehensive correlation analysis of key academic variables using both a pairs matrix and a heatmap.

```{r}
#| label: correlation-heatmap
#| fig-width: 9
#| fig-height: 7

df_var <- df %>%
  select(attend, termGPA, priGPA, ACT, final, atndrte, hwrte)

ggcorr(df_var, label = TRUE, label_round = 2, hjust = 0.6, size = 4,
       layout.exp = 2) +
  labs(title = "Correlation Heatmap of Academic Variables") +
  theme(plot.title = element_text(face = "bold", size = 14))
```

**Key Findings:**

- **Strongest correlation:** `attend` and `atndrte` (r = 1.0) — these are directly derived from each other
- **Next strongest:** `termGPA` and `priGPA` (r ≈ 0.7) — strong academic consistency
- **Attendance vs. homework for GPA:** Attendance rate shows stronger correlations with both term and prior GPA than homework submission, suggesting that class attendance may be a more effective strategy for improving grades

## Interactive Visualization

Using `plotly` to create an interactive scatter plot that allows exploration of the relationship between prior GPA and term GPA.

```{r}
#| label: interactive-plot
#| fig-width: 10
#| fig-height: 6

df_soph_labeled <- df %>%
  mutate(soph_label = factor(soph, levels = c(0, 1),
                              labels = c("Not Sophomore", "Sophomore")))

scatter_plt <- ggplot(df_soph_labeled, aes(x = priGPA, y = termGPA, color = soph_label)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c("Not Sophomore" = "#e74c3c", "Sophomore" = "#2c3e50")) +
  labs(
    title = "Prior GPA vs. Term GPA",
    subtitle = "Interactive — hover to explore individual data points",
    x = "Prior GPA", y = "Term GPA", color = "Status"
  ) +
  theme_minimal(base_size = 13)

ggplotly(scatter_plt, tooltip = c("x", "y", "color"))
```

A strong positive relationship exists between prior and current GPA across both class levels, demonstrating academic consistency regardless of sophomore status.
