---
title: "R Programming & Functional Programming"
author: "Isabelle Garza"
description: "Control flow, custom functions, apply/map patterns, and real-world data analysis in R."
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    code-line-numbers: true
execute:
  warning: false
  error: true
  freeze: auto
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(purrr)
library(gt)
```

This page showcases R programming concepts including conditional logic, loops, custom functions, the `apply` family, `purrr::map` functions, and their application to real-world data analysis.

## Control Flow: If/Else Logic

### Sales Performance Classifier

```{r}
#| label: sales-classifier

sales <- 25

if (sales < 20) {
  print("Not enough for today.")
} else if (sales >= 20 && sales <= 30) {
  print("Average Day")
} else {
  print("You had a great day!")
}
```

### Tax Rate Calculator

A function-style approach using `ifelse()` to apply category-based tax rates:

```{r}
#| label: tax-calculator

category <- "A"
price <- 100

tax_rate <- ifelse(category == "A", 0.08,
                   ifelse(category == "B", 0.10, 0.20))

total <- price + price * tax_rate
message(paste("A tax rate of", tax_rate * 100, "% is applied. The total price is", total))
```

## Loops: Counting and Filtering

### Counting Even Numbers

```{r}
#| label: count-evens

q4 <- c(2, 5, 3, 9, 8, 11, 6)
count <- 0

for (i in q4) {
  if (i %% 2 == 0) {
    count <- count + 1
  }
}

paste("Even numbers found:", count)
```

### Conditional Printing with Loops

```{r}
#| label: conditional-loop

q5 <- 1:8
for (i in seq_along(q5)) {
  val <- q5[i]
  if (val > 4 && val %% 2 == 0) {
    print(val)
  } else {
    print("Condition not satisfied")
  }
}
```

## Custom Functions

### Power Function

```{r}
#| label: power-function

pow <- function(x, y) {
  result <- x^y
  paste0(x, " raised to the power of ", y, " is ", result, ".")
}

pow(8, 2)
pow(3, 4)
pow(2, 10)
```

## Apply Family vs. Map Family

Comparing base R's `apply`/`lapply`/`sapply` functions with `purrr`'s `map` functions for functional programming.

### Row-Level Statistics with `apply()`

```{r}
#| label: apply-demo

df <- data.frame(x = 1:4, y = 5:8, z = 10:13)
df$mean <- apply(df, 1, mean)
df$sum  <- apply(df, 1, sum)

df %>%
  gt() %>%
  tab_header(title = "Row-Level Statistics Using apply()")
```

### Square Roots: lapply vs. map

```{r}
#| label: lapply-vs-map

q8 <- c(12, 18, 6)

# apply family
cat("lapply result:\n")
lapply(q8, sqrt)

# map family
cat("\nmap result:\n")
map(q8, sqrt)
```

> **Key difference:** `lapply()` always outputs a list, while `sapply()` tries to simplify. In `purrr`, `map()` returns a list, while `map_dbl()` returns a typed numeric vector — making results easier to work with downstream.

## Student-Teacher Ratios: Functional Approach

Using custom functions and `lapply()` to automate chart generation across multiple years of data.

```{r}
#| label: student-teacher-data
#| include: false

student_ratio <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-07/student_teacher_ratio.csv")
```

### Reusable Chart Function

```{r}
#| label: ratio-function

plot_ratio <- function(year_input) {
  student_ratio %>%
    filter(year == year_input) %>%
    top_n(10, student_ratio) %>%
    ggplot(aes(x = reorder(country, student_ratio), y = student_ratio)) +
    geom_col(fill = "#3498db", alpha = 0.85) +
    coord_flip() +
    labs(
      title = "Top 10 Countries by Student-Teacher Ratio",
      subtitle = paste("Year:", year_input),
      x = "", y = "Student-Teacher Ratio"
    ) +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(face = "bold"))
}
```

### Generated Charts: 2012–2017

::: {.panel-tabset}

#### 2012

```{r}
#| label: ratio-2012
plot_ratio(2012)
```

#### 2013

```{r}
#| label: ratio-2013
plot_ratio(2013)
```

#### 2014

```{r}
#| label: ratio-2014
plot_ratio(2014)
```

#### 2015

```{r}
#| label: ratio-2015
plot_ratio(2015)
```

#### 2016

```{r}
#| label: ratio-2016
plot_ratio(2016)
```

#### 2017

```{r}
#| label: ratio-2017
plot_ratio(2017)
```

:::

## Nested Data & Map: Ozone by Month

Using nested data frames and `map_dbl()` to calculate average ozone by month — a functional programming pattern that avoids repetitive code.

```{r}
#| label: ozone-nested

calc_mean <- function(data) {
  mean(data$Ozone, na.rm = TRUE)
}

ozone_by_month <- function(df) {
  df %>%
    group_by(Month) %>%
    nest() %>%
    mutate(mean_ozone = map_dbl(data, calc_mean)) %>%
    select(Month, mean_ozone) %>%
    arrange(Month)
}

result <- ozone_by_month(airquality)

result %>%
  mutate(Month = recode(Month,
    `5` = "May", `6` = "June", `7` = "July",
    `8` = "August", `9` = "September"
  )) %>%
  gt() %>%
  tab_header(
    title = "Average Ozone by Month",
    subtitle = "Computed using nested data frames and map_dbl()"
  ) %>%
  cols_label(Month = "Month", mean_ozone = "Mean Ozone (ppb)") %>%
  fmt_number(columns = mean_ozone, decimals = 1)
```

```{r}
#| label: ozone-viz
#| fig-width: 8
#| fig-height: 5

result %>%
  mutate(Month_name = recode(Month,
    `5` = "May", `6` = "June", `7` = "July",
    `8` = "August", `9` = "September"
  )) %>%
  ggplot(aes(x = fct_reorder(Month_name, mean_ozone), y = mean_ozone)) +
  geom_col(fill = "#e67e22", alpha = 0.85) +
  geom_text(aes(label = round(mean_ozone, 1)), hjust = -0.2, fontface = "bold") +
  coord_flip() +
  labs(
    title = "Average Ozone Levels by Month",
    subtitle = "July and August have the highest ozone concentrations",
    x = "", y = "Mean Ozone (ppb)"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold")) +
  expand_limits(y = max(result$mean_ozone) * 1.15)
```

**Key takeaway:** Functional programming in R — using nested data, custom functions, and `map()` — eliminates repetitive code and makes analyses reproducible and scalable.
